---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    toc: yes
  pdf_document: default
  word_document: default
---

------------------------------------

## Loading and preprocessing the data
```{r, LoadDataProcessData}
strUrl <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
strFile <-"activity.zip"
if (!file.exists(strFile))
{
     download.file(url = strUrl, destfile = "activity.zip", method = "curl", mode = "wb")  
}
lstOfFiles <-unzip(zipfile = strFile)
#
df <-read.csv(file = lstOfFiles[1], colClasses = c("integer","character","integer"))
#
str(df)
#
df$interval.ori <-df$interval
df$interval <-sprintf(fmt = "%04d", df$interval)
#
df$date.ori <-df$date
df$date <-as.Date(df$date)
#
df$dayofweek <-weekdays(df$date)
#
df$daytype <-"weekday"
df[substr(df$dayofweek,1,1) == "S", "daytype"] <-"weekend"
df$daytype <-factor(df$daytype, levels = c("weekday","weekend"))
#
str(df)
```

* #### The data was loaded using `read.csv`.
* #### Existing variables were copied and then transformed: `interval.ori` & `date.ori`.
* #### New variables were created: `dayofweek` & `daytype`.

----------------------------------------------------

## What is the mean total number of steps taken per day?
```{r, TotalStepsPerDay}
# 1. Calculate the total number of steps taken per day
TotalStepsPerDay <-tapply(X = df$steps, INDEX = df$date, FUN = sum, na.rm = TRUE)
head(TotalStepsPerDay, n = 25)
```

```{r, Plot_TotalStepsPerDay}
# 2. Make a histogram of the total number of steps taken each day.
hist(x = TotalStepsPerDay, breaks = 60, xlab = "Total Steps Per Day - NAs removed")
# 3. Calculae and report the mean and the median of the total number of steps taken per day.
(meanTotalStepsPerDay <-mean(TotalStepsPerDay))
(medianTotalStepsPerDay <-median(TotalStepsPerDay))
```
* #### The mean number of steps taken each day is `r round(meanTotalStepsPerDay, digits = 0)`.
* #### The median number of steps taken each day is `r medianTotalStepsPerDay`.

----------------------------------------------

## What is the average daily activity pattern?
```{r, AvgDailyActivityPattern}
AvgStepsPerInterval <-tapply(X = df$steps, INDEX = df$interval, FUN = mean, na.rm = TRUE)
head(AvgStepsPerInterval, n = 25)
# 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```

```{r, Plot_AvgStepsPerInterval}
plot(names(AvgStepsPerInterval), AvgStepsPerInterval,type = "l", xlab = "5 Minute Intervals")
# 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
(maxAvgStepsPerInterval <-AvgStepsPerInterval[AvgStepsPerInterval == max(AvgStepsPerInterval)])
```

* #### The average daily activity pattern was plotted.
* #### The 5-minute interval that, on average, contains the maximum number of steps (`r round(maxAvgStepsPerInterval, digits = 0)`) is interval `r names(maxAvgStepsPerInterval)`.

--------------------------

## Imputing Missing Values.
```{r, ImputMissingValues}
# 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
(missingcases <-sum(!complete.cases(df)))
(missingsteps <-sum(is.na(df$steps)))
# 2. Devise a strategy for filling in all of the missing values in the dataset.
# create a matrix of avg steps per interval (row name) by day of week (column name)
mxAvgStepsPerIntervalByDow <-tapply(X = df$steps, INDEX = list(df$interval,df$dayofweek), mean, na.rm = TRUE)
# 3. Create a new dataset that is equal to the original dataset but with missing data filled in.
df_narpl <-df
# make copy of original steps variable
df_narpl$steps.ori <-df_narpl$steps
# loop through data frame "df_narpl" and replace NA steps with avg steps from replacement matrix
for (i in 1:nrow(df_narpl))
{
     isteps <-df_narpl[i,"steps"]
     iinterval <-df_narpl[i,"interval"]
     idayofweek <-df_narpl[i,"dayofweek"]
     # if steps is missing, replace with avg from dfcmatrix
     if(is.na(isteps))
     {
          df_narpl[i,"steps"] <-mxAvgStepsPerIntervalByDow[iinterval,idayofweek]
     }
}
# 4. Make a histogram of the total number of steps taken each day and calculate the mean and median total number of steps taken per day.
TotalStepsPerDay_narpl <-tapply(X = df_narpl$steps, INDEX = df_narpl$date, FUN = sum, na.rm = TRUE)
```
* #### The total number of missing values in the dataset is `r missingsteps`.
* #### A new data frame `df_narpl` was created where missing steps (NAs) were replaced with the average steps of the matching interval and day of the week. Parts of the update matrix `mxAvgStepsPerIntervalByDow` are shown below:

```{r, ShowUpdateMatrix}
head(mxAvgStepsPerIntervalByDow)
head(mxAvgStepsPerIntervalByDow[97:117,],12)
tail(mxAvgStepsPerIntervalByDow)
```

```{r, Plot_TotalStepsPerDayNAsReplaced}
# histogram
hist(x = TotalStepsPerDay_narpl, breaks = 60, xlab = "Total Steps Per Day - NAs Replaced with Avg Values")
# mean and median
(meanTotalStepsPerDay_narpl <-mean(TotalStepsPerDay_narpl))
(medianTotalStepsPerDay_narpl <-median(TotalStepsPerDay_narpl))
# compare with mean where incomplete records (NAs) were removed
(meanDiff <-meanTotalStepsPerDay_narpl - meanTotalStepsPerDay)
# compare with median where incomplete records (NAs) were removed
(medianDiff <-medianTotalStepsPerDay_narpl - medianTotalStepsPerDay)
```

* #### The impact of imputing the missing data (replacing NAs with the average steps per mathcing interval and day of the week) changed the histogram of the total number of steps per day. The mean number of steps per day increased by `r round(meanDiff, digits = 0)` steps and the median number of steps per day increased by `r round(medianDiff, digits = 0)` steps.

----------------------------------------------------------------------------

## Are there differences in activity patterns between weekdays and weekends?
```{r, ActivityWeekdayVsWeekend}
df_narpl$dayofweek <-weekdays(df_narpl$date)
# 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or a weekend day.
df_narpl$daytype <-"weekday"
df_narpl[substr(df_narpl$dayofweek,1,1) == "S", "daytype"] <-"weekend"
df_narpl$daytype <-factor(df_narpl$daytype)
# 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekdays or weekend days (y-axis)
# create a matrix of average number of steps taken per interval per daytype (weekday & weekend)
mxAvgStepsPerIntervalByDayType <-tapply(X = df_narpl$steps, INDEX = list(df_narpl$interval, df_narpl$daytype), FUN = mean, na.rm = TRUE)
str(mxAvgStepsPerIntervalByDayType)
# reshape (melt) the matrix to a data frame with 3 columns : interval, day type and avg steps per interval
require(reshape2)
df_mlt <-melt(data = mxAvgStepsPerIntervalByDayType, measure.vars = c("weekday","weekend"), id.vars = c("interval"))
# update column/variable names
names(df_mlt) <-c("timeInterval","dayType","avgSteps")
str(df_mlt)
# create panel plots of avg steps by interal for each day type (weekday & weekend)
```

```{r, Plot_WeekdayVsWeekendActivity}
require("lattice")
xyplot(avgSteps ~ timeInterval | dayType, data = df_mlt, type = "l", layout = c(1,2), xlab = "5 Minute Intervals", ylab = "Avg Steps" )
```

* #### The panel plots show a different pattern of interval activity for weekdays vs weekends --- one prominent difference is the increased activity between 10am and 6pm on weekends compared to the same period on weekdays.

-----------------------------------------------